Highly Available Transactions: Virtues and Limitations

### 2013-04-29 writeup/evaluation.tex
eval rewrite

To better understand the performance implications of HAT guarantees in
a real-world environment and to validate the latency-based analysis in
Section~\ref{sec:latency}, we deployed a distributed database
prototype on public cloud infrastructure. We compared the behavior of
eventual consistency, per-item mastering (wherein all operations to a
given item required contacting a designated master server), Read
Committed, and TA with read your writes guarantees. We deployed an
experimental Java-based distributed datastore backed by LevelDB both
within and across EC2 regions and ran the YCSB benchmark~\cite{ycsb}
to determine the impact of latency. We partitioned the database into
two fully replicated clusters of three \texttt{m1.xlarge} instances
each. For HAT, we stuck YCSB clients to their local cluster and, to
provide transactional functionality, grouped every six YCSB operations
from the default workload to form a transaction. We report the median
of five trials per experiment.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=.8\columnwidth]{figs/strategylegend.pdf}
\end{center}\vspace{-2em}
\begin{center}\small\textbf{Within \texttt{us-east (VA)}}\end{center}\vspace{-1.5em}
\includegraphics[width=0.90\columnwidth]{figs/finals/2lan-threads-lats-log.pdf}\vspace{-.5em}
\includegraphics[width=0.90\columnwidth]{figs/finals/2lan-threads-thru.pdf}
\begin{center}\small\textbf{Between \texttt{us-east (CA)} and \texttt{us-west-2 (OR)}}\end{center}\vspace{-1.5em}
\includegraphics[width=0.90\columnwidth]{figs/finals/2wan-threads-lats-log.pdf}\vspace{-.5em}
\includegraphics[width=0.90\columnwidth]{figs/finals/2wan-threads-thru.pdf}
\begin{center}\small Between \textbf{\texttt{us-east} \texttt{(VA)}, \texttt{us-west-1} \texttt{(CA)},\\ \texttt{us-west-2} \texttt{(OR)}, \texttt{eu-west} \texttt{(IR)}, \texttt{ap-northeast} \texttt{(SI)}}\end{center}\vspace{-1.5em}
\includegraphics[width=0.90\columnwidth]{figs/finals/5wan-threads-lats-log.pdf}\vspace{--.5em}
\includegraphics[width=0.90\columnwidth]{figs/finals/5wan-threads-thru.pdf}
\caption{YCSB performance for two clusters of five servers each
  deployed within a single datacenter and cross-datacenters.}
\label{fig:wan-exp}
\end{figure}

\begin{figure}[t!]
\begin{center}
\includegraphics[width=0.90\columnwidth]{figs/finals/scaleout-thru.pdf}
%\includegraphics[width=0.90\columnwidth]{figs/finals/scaleout-thru-perserver.pdf}
\end{center}\vspace{-2em}
\caption{Scale-out of TA, Eventual, and RC.}
\label{fig:scaleout}
\end{figure}

\begin{figure}[t!]
\begin{center}
\includegraphics[width=0.90\columnwidth]{figs/finals/txnlen-thru.pdf}
\end{center}\vspace{-2em}
\caption{Transaction length versus throughput.}
\label{fig:txlen}
\end{figure}

\begin{figure}[t!]
\begin{center}
\includegraphics[width=0.90\columnwidth]{figs/finals/wprop-thru.pdf}
\end{center}\vspace{-2em}
\caption{Proportion of reads and writes versus throughput.}
\label{fig:rprop}
\end{figure}

As Figure~\ref{fig:wan-exp} shows, within a single datacenter, a
mastered implementation achieves approximately double the latency and
slightly over half of the throughput of a HAT implementation. This is
because, given two replicas for each key, the mastered implementation
serves requests from only one. The mastered implementation still
achieves $19,797$ transactions per second---over 6,000 per
master---and is only $15.56$ms slower at peak compared to eventual
consistency. However, eventual consistency and HAT models achieve
substantially higher throughput--$37,575$ and $36,178$ transactions
per second, respectively--and lower latency---$48.3\%$ and $47.2\%$
respectively. This difference is unsurprising because, in these weaker
models, all servers can serve both reads and writes, effectively
doubling the capacity of the system. The weaker models still propagate
new values between replicas, but this process is asynchronous and,
aside from consuming disk I/O and network bandwidth, does not
substantially impact clients. As expected, HAT models perform
similarly to eventual consistency: our Read Committed implementation
buffers client writes and adds no additional server-side overhead,
while TA requires additional metadata with each write but has less
than 5\% performance impact compared to eventual consistency.

In comparison, over multiple datacenters, the cost of coordination
increases. The HAT and eventual consistency implementations continue
to enjoy local RTTs only, and, with the exception of a slight
performance degradation in TA---which we believe is attributable to
Java garbage collection overheads in TA's inbound anti-entropy
buffers---performance remains unchanged. However, each YCSB update to
a remote replica ($\frac{2}{3}$ in one datacenter, $\frac{1}{3}$ in
the other) of requiring a remote RTT between Virginia and Oregon (or
vice-versa) to reach the master. This leads to a latency increase of
around $166$ms ($746\%$) per transaction compared to a
single-datacenter deployment. In order to attain reasonable throughput
for the mastered deployment, we had to greatly increase the number of
concurrent clients accessing the database, and we found that threading
and socket overheads quickly overwhelmed the database server. Over a
wide-area network, the increased cost of RTTs dramatically affects
system efficiency.

This brief set of experiments is not meant as an exhaustive study, but
it validates our earlier intuition about the cost of violating
HAT-compliance. As Deutsch points out, ignoring factors such as
latency can ``cause big trouble and painful learning
experiences''~\cite{fallacies-deutsch}---in a single-site context,
paying the cost of coordination may be tenable, but, especially as
services are geo-replicated,  costs increase.

>>>>

To further understand the performance implications of HAT guarantees
in a real-world environment, we implemented a HAT database
prototype. Across a range of deployments on public cloud
infrastructures, we found that, as Section~\ref{sec:latency}'s
measurements suggested, ``strongly consistent'' algorithms incur
substantial latency penalties: from 10 to 100 times their HAT
counterparts. Moreover, HAT algorithms like TA---while more
resource-intensive than basic eventual consistency---are able to scale
well to many servers.

\vspace{.5em}\noindent\textit{Implementation.} Our prototype database
is a partially replicated (hash-based, partitioned) key-value backed
by LevelDB and implemented in Java using Apache Thrift. Our prototype
currently supports eventual consistency (via simple anti-entropy) and
HAT RU, RC, and TA guarantees (as described in
Section~\ref{sec:hats}). It also supports non-HAT operation whereby
all operations for a given key are routed to a (randomly) designated
\textit{master} replica for each key (i.e., PNUTS~\cite{pnuts}) as
well as distributed two-phase locking. Servers are durable: they
synchronously write to LevelDB before responding to client requests,
while pending writes in TA are sent to a write-ahead log, which is
synchronously flushed to disk.

\vspace{.5em}\noindent\textit{Configuration.} We deploy the database
in \textit{clusters}---sets of database servers that contain a single,
fully replicated copy of the data---typically across datacenters,
sticking all clients within a datacenter to their respective
cluster. By default, we use 5 Amazon EC2 \texttt{m1.xlarge} instances
as servers in each cluster. For our workload, we link our client
library to the YCSB benchmark~\cite{ycsb}, grouping every eight YCSB
operations from the default workload (50\% reads, 50\% writes) to form
a transaction. We increase the number of keys in the workload from the
default $1,000$ to $100,000$ with uniform random key access, keeping
the default value size of $1KB$, and running YCSB for 180 seconds per
configuration. Clients trivially achieve read-your-writes and
monotonic reads guarantees via stickiness.

\begin{figure}[t!]
\begin{center}
\hspace{2em}\includegraphics[width=.8\columnwidth]{figs/strategylegend.pdf}
\end{center}\vspace{-2em}
\begin{center}\small\textbf{A.) Within \texttt{us-east} \texttt{(VA)}}\end{center}\vspace{-1.5em}
\includegraphics[width=\figfactor\columnwidth]{figs/finals/2lan-threads-lats.pdf}\vspace{-.5em}
\includegraphics[width=\figfactor\columnwidth]{figs/finals/2lan-threads-thru.pdf}\vspace{-.75em}
\begin{center}\small\textbf{B.) Between \texttt{us-east} \texttt{(CA)} and \texttt{us-west-2} \texttt{(OR)}}\end{center}\vspace{-1.5em}
\includegraphics[width=\figfactor\columnwidth]{figs/finals/2wan-threads-lats-log.pdf}\vspace{-.5em}
\includegraphics[width=\figfactor\columnwidth]{figs/finals/2wan-threads-thru.pdf}\vspace{-.75em}
\begin{center}\small \textbf{C.) Between}{ \texttt{us-east} \texttt{(VA)}, \texttt{us-west-1} \texttt{(CA)},\\ \texttt{us-west-2} \texttt{(OR)}, \texttt{eu-west} \texttt{(IR)}, \texttt{ap-northeast} \texttt{(SI)}}\end{center}\vspace{-1.5em}
\includegraphics[width=\figfactor\columnwidth]{figs/finals/5wan-threads-lats-log.pdf}\vspace{--.5em}
\includegraphics[width=\figfactor\columnwidth]{figs/finals/5wan-threads-thru.pdf}
\caption{YCSB performance for two clusters of five servers each
  deployed within a single datacenter and cross-datacenters.}
\label{fig:wan-exp}
\end{figure}

\vspace{.5em}\noindent\textit{Geo-replication.} We first deploy the
database prototype across an increasing number of
datacenters. Figure~\ref{fig:wan-exp}A shows that, when operating two
clusters within a single datacenter, employing a single master for
each data item results in approximately half the throughput and double
the latency of an eventually consistent configuration: the HAT models
are able to utilize double the number of servers. RC---essentially
eventual consistency with buffering---is almost identical to eventual
consistency, while TA---which incurs two writes for every client-side
write (i.e., pending writes are sent to a WAL then subsequently moved
into LevelDB once stable)---achieves approximately 75\% of the
throughput. Latency increases linearly with the number of YCSB clients
due to contention within LevelDB.

 In contrast, when the two clusters are deployed across the
 continental United States (Figure~\ref{fig:wan-exp}B), the average
 latency of mastered operation increases to $300$ms (a $278$--$4257\%$
 latency increase; average $37$ms latency per operation). For the same
 number of YCSB client threads, mastered operation has substantially
 lower throughput than the HAT configurations. Increasing the number of YCSB
 clients \textit{does} increase the throughput of mastered operation,
 but our Thrift-based connection processing did not gracefully handle
 more than several thousand concurrent connections. In contrast,
 across two datacenters, the performance of eventual, RC, and TA is
 near-identical to a single-datacenter deployment.

When we deployed five clusters (as opposed to two, as before) across
the five AWS datacenters with lowest communication cost
(Figure~\ref{fig:wan-exp}C), the trend continues: mastered latency
increases to nearly $800$ms per transaction. As an attempt at reducing
this overhead, we implemented and benchmarked a variant of
quorum-based replication as in Dynamo~\cite{dynamo}, where clients
sent requests to all replicas, which completed as soon as a majority
of servers responded (guaranteeing regular
semantics~\cite{herlihy-art}); this strategy (not pictured) did not
substantially improve performance due to the network topology and
because worst-case server load was unaffected. In this configuration,
TA's relative throughput decreased: every YCSB \texttt{put} operation
resulted in four \texttt{put} operations on remote replicas and,
accordingly, the cost of anti-entropy increased (e.g., each server
processed four incoming anti-entropy streams---as opposed to one
before---reducing the opportunity for batching and decreasing
available resources for for incoming client requests). This in turn
increased garbage collection activity and, more importantly, IOPS when
compared to Eventual and RC, causing throughput to peak at around half
of Eventual. With an in-memory persistence (i.e., no LevelDB or WAN),
TA throughput was within 20\% of eventual.

We have intentionally omitted performance data for two-phase
locking. Mastered configuration performed \textit{far} better than our
textbook implementation, which, in addition to requiring a WAN
round-trip per transaction, also incurred substantial overheads due to
mutual exclusion via locking. We expect that, while techniques like
those recently proposed in Calvin~\cite{calvin} can reduce the
overhead of serializable transactions by avoiding locking, our
master-based implementation and the data from
Section~\ref{sec:latency} are reasonable lower bounds on latency.

\begin{figure}[t!]
\begin{center}
\hspace{2em}\includegraphics[width=.8\columnwidth]{figs/strategylegend.pdf}\vspace{-2em}
\end{center}
\begin{center}
\includegraphics[width=\figfactor\columnwidth]{figs/finals/txnlen-thru.pdf}
\end{center}\vspace{-2.25em}
\caption{Transaction length versus throughput.}\vspace{-.5em}
\label{fig:txlen}
\end{figure}
\begin{figure}[t!]
\begin{center}
\includegraphics[width=\figfactor\columnwidth]{figs/finals/wprop-thru.pdf}
\end{center}\vspace{-2.25em}
\caption{Proportion of reads and writes versus throughput.}\vspace{-.5em}
\label{fig:rprop}
\end{figure}
\begin{figure}[t!]
\begin{center}
\includegraphics[width=\figfactor\columnwidth]{figs/finals/scaleout-thru.pdf}
%\includegraphics[width=\figfactor\columnwidth]{figs/finals/scaleout-thru-perserver.pdf}
\end{center}\vspace{-2.25em}
\caption{Scale-out of TA, Eventual, and RC.}
\label{fig:scaleout}
\end{figure}

\vspace{.5em}\noindent\textit{Transaction length.} As shown in
Figure~\ref{fig:txlen} (clusters in Virginia and Oregon), throughput
of eventual consistency, RC, and mastered operation are unaffected by
transaction length. In contrast, TA's throughput decreases linearly
with increased transaction length: with $1$ operation per transaction,
TA throughput is within 18\% of eventual ($34$ bytes overhead), and
with $128$ operations per transaction, TA throughput is within $60\%$
($1898$ bytes overhead). This is primarily because our TA algorithm
requires metadata proportional to the length of each transaction,
consuming IOPS and network bandwidth.

\vspace{.5em}\noindent\textit{Read proportion.} The default equal
proportion of reads and writes is fairly pessimistic: for example,
Facebook reports $99.8\%$ reads for their workload~\cite{eiger}. As
shown in Figure~\ref{fig:rprop} (clusters in Virginia and Oregon),
with all reads, TA is within $4.8\%$ of eventual; with all writes,
eventual throughput decreases by $288.8\%$ compared to all reads and
TA is within $33\%$ of eventual. At $99.8\%$ reads, TA incurs a $7\%$
overhead ($5.8\%$ for in-memory storage).

\vspace{.5em}\noindent\textit{Scale-out.} One of the key benefits of
our HAT algorithms is that they are shared-nothing, meaning they
should not compromise scalability. Figure~\ref{fig:scaleout} shows
that varying the number of servers across two clusters in Virginia and
Oregon (with $15$ YCSB clients per server) results in linear scale-out
for both eventual, RC, and TA. RC and eventual scale perfectly with an
approximately $5$x throughput increase moving from $5$ to $25$ servers
per cluster. For the same configuration, TA scales by $3.8$x,
achieving over $260,000$ operations per second. We believe that TA's
performance is due to resource contention---with a memory-backed
database (instead of LevelDB), TA scales by $4.25$x (not shown)---and
increased garbage collection activity.

\vspace{.5em}\noindent\textit{Summary.} Our experimental prototype
confirms our earlier analytical intuitions. HAT systems can provide
useful semantics without substantial performance penalties. In
particular, our TA algorithm can achieve throughput competitive with
eventual consistency at the expense of increased disk and network
utilization. Perhaps more importantly, all HAT algorithms circumvent
high WAN latencies inevitable with non-HAT implementations. Our results
highlight Deutcsh's observation that ignoring factors such as latency
can ``cause big trouble and painful learning
experiences''~\cite{fallacies-deutsch}---in a single-site context,
paying the cost of coordination may be tenable, but, especially as
services are geo-replicated, costs increase.

### 2013-04-29 writeup/hat-analysis.tex
eval rewrite

\end{tikzpicture}
\label{fig:hat-order}
\caption{HAT, sticky HAT (in boxes, blue), and unavailable models
  (circled, red) from Table~\protect\ref{table:hatcompared} compared
  graphically. Directed edges represent ordering by model
  strength. Models that do not share a common ancestor can be
  simultaneously achieved, and the resulting availability is that of
  the weakest available model in the combination.}
\label{fig:hatcompared}
\end{figure}

>>>>

\end{tikzpicture}
\label{fig:hat-order}
\caption{Partial ordering of HAT, sticky HAT (in boxes, blue), and
  unavailable models (circled, red) from
  Table~\protect\ref{table:hatcompared}. Directed edges represent
  ordering by model strength. Incomparable models can be
  simultaneously achieved, and the availability of a combination of
  models has the availability of the least available individual
  model.}
\label{fig:hatcompared}
\end{figure}

### 2013-04-29 writeup/intro.tex
some graphs and joe comments
eval rewrite

Our main contributions in this paper are as follows. We fill the gap
between the plethora of previously proposed isolation and consistency
models and the goal of high availability.  We classify which among the
wide array of models are achievable with high availability, denoting
them as {\em Highly Available Transactions} (HATs). In doing so, we
demonstrate that although many implementations of HAT semantics are
not highly available, this is an artifact of the implementations
rather than an inherent property of the semantics. Our investigation
shows that, besides serializability, only Snapshot Isolation and
Repeatable Reads are not HAT-compliant, while most other isolation
levels are achievable with high availability. We also demonstrate that
many weak data consistency models from distributed systems are also
HAT-compliant.

Our investigation is based on both impossibility results and several
constructive proof-of-concept algorithms. For example, Snapshot
Isolation (SI) and Repeatable Reads (RR) are not HAT-compliant because
they require detecting conflicts between concurrent updates (as needed
for preventing Lost Updates or Write Skew phenomena), which we prove
is unavailable. However, Read Committed isolation, Transactional
Atomicity, and many of the weaker data consistency models from
database and distributed systems are achievable via novel algorithms
that rely on multi-versioning and limited client-side caching. For
several guarantees, such as causal consistency with phantom prevention
and ANSI Repeatable Read, we consider a modified form of high
availability in which clients ``stick to'' (i.e., have affinity with)
at least one server (a property which is often implicit in the
distributed systems literature~\cite{herlihy-art, eiger, cac} but
which requires explicit consideration in a client-server replicated
database context). This sticky availability is easily implementable
and widely employed~\cite{eiger, vogels-defs} but is strictly less
available than traditional definitions of high availability.

>>>>

Our main contributions in this paper are as follows. We fill the gap
between the plethora of previously proposed isolation and consistency
models and the goal of high availability.  We classify which among the
wide array of models are achievable with high availability, denoting
them as {\em Highly Available Transactions} (HATs). In doing so, we
demonstrate that although many implementations of HAT semantics are
not highly available, this is an artifact of the implementations
rather than an inherent property of the semantics. Our investigation
shows that, besides serializability, only Snapshot Isolation and
Repeatable Reads are not HAT-compliant, while most other isolation
levels are achievable with high availability. We also demonstrate that
many weak data consistency models from distributed systems are also
both HAT-compliant and simultaneously achievable with traditional ACID
properties.

Our investigation is based on both impossibility results and several
constructive proof-of-concept algorithms. For example, Snapshot
Isolation (SI) and Repeatable Reads (RR) are not HAT-compliant because
they require detecting conflicts between concurrent updates (as needed
for preventing Lost Updates or Write Skew phenomena), which we prove
is unavailable. However, Read Committed isolation, Transactional
Atomicity, and many of the weaker data consistency models from
database and distributed systems are achievable via novel algorithms
that rely on multi-versioning and limited client-side caching. For
several guarantees, such as causal consistency with phantom prevention
and ANSI Repeatable Read, we consider a modified form of high
availability in which clients ``stick to'' (i.e., have affinity with)
at least one server (a property which is often implicit in the
distributed systems literature~\cite{herlihy-art, eiger, cac} but
which requires explicit consideration in a client-server replicated
database context). This sticky availability is easily implementable
and widely employed~\cite{eiger, vogels-defs} but is more restrictive
than traditional definitions of high availability.
